---
title: "Project proposal"
author: "Gold Analysts - Mekkawy, Nissan, Razvan"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
library(dplyr)
library(skimr)
```

## 1. Intro 

We are developing our research based on a database provided by Promise Early Education Programs: Early Head Start and Head Start. They focus on children’s developmental outcomes. 

The data is collected using the Teaching Strategies Gold (TS GOLD) assessment, which analyzes growth towards Widely Held Expectations (WHE) for typically developing children at each stage. 

This collection method categorizes children’s performance in every vital classification: Physical Examination, Language, Cognitive, Social-Emotional, Literacy and Mathematics.

We plan on researching how COVID-19 impacted different groups of students. We will analyze these groups in two categories: based on their age or based on whether they have an Individualized Educational Program (IEP) or not.

The dataset used for this analysis contains 180 observations and 10 variables in the main structured table (prof_2018_2019). These variables are `Category`, `WHE Bottom`, `WHE Top`, `# Children`, `Average`, `# Below`, `% Below`, `# Meeting / Exceeding`, `% Meeting / Exceeding and Time Period`.


## 2. Data

We are currently tidying up the 6 datasets we plan on using for our research. Each of them presents specific data collected in a school year (e.g. 2018-2019, 2023-2024).

Meanwhile, since the datasets from the NGO have the same variables, we will use a past dataset to explain the structure. Our variables are: 
- `Bottom`: Lowest Score (for agency/ classroom)
- `Top`: Highest Score (for agency/ classroom
- `# Children`: Total # of children assessed
- `Average` : Average Score of children (for agency/ classroom)
- `# Below` :# of children below developmental expectations
- `% Below` : % of children below developmental expectations

```{r reading-data}
prof_2018_2019 <- read_excel("../data/ignore/file_show.xlsx")
```

```{r visualizing-data}
glimpse(prof_2018_2019)
```

```{r skimming-data}
skim(prof_2018_2019)
```

## 3. Data analysis plan

We will analyze our data from the following perspective:
- Pre-COVID-19: `2018_2019`
- COVID-19: `2020_2021` and `2021_2022`
- Post-COVID-19: `2022_2023` , `2023_2024` and `2024_2025`

For our research questions, we will analyze the overall performance in each general academic category: `Cognitive`, `Language`, `Literacy`, `Mathematics`, `Physical`, and `Social-Emotional`. 

For an in-depth categorization of our data, we will divide the data based on `IEP` and `Age`. 

To compare student success between these groups, we decided to create a new variable representing the percentage of students who meet or exceed expectations. Since the data is divided into semesters, we will combine the answers into one year. We will create a new variable that will store that number. 

#### Graphs:

1. Ridgeline Density Plot - General:  We aim to show how the pandemic affected students overall. To visualize this, we plan to use a density ridgeline graph, where the x-axis represents the school year and the y-axis represents the percentage of children meeting or exceeding expectations. We also plan to separate the data by category using facets for clearer comparisons.

2. Facet Line Graph - Age Groups: We will specifically focus on how the COVID-19 pandemic impacted different age groups, ranging from 0 to 5. We plan to use a line graph, where the x-axis represents the school year and the y-axis represents the percentage of students who meet or exceed expectations. The graph will be faceted by category, and different line colors will represent each age group.

3. Facet Density Plot - IEP Groups: We will use similar axes (x = school year, y = percentage of students who meet or exceed expectations) and the same facet categories as in our second graph. However, instead of focusing on different age groups, this graph will compare students with IEPs to those without IEPs to determine whether children with IEPs were more affected by the pandemic. For now, this graph will include data only up to Winter 2023, since we do not yet have IEP data for the 2023–2024 and 2024–2025 academic years. We are currently waiting for that information from Monica.
