---
title: "Project proposal"
author: "Gold Analysts - Mekkawy, Nissan, Razvan"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
library(dplyr)
library(skimr)
```

## 1. Intro 

We are developing our research based on a database provided by Promise Early Education Programs: Early Head Start and Head Start. They focus on children’s developmental outcomes. 

The data is collected using the Teaching Strategies Gold (TS GOLD) assessment, which analyzes growth towards Widely Held Expectations (WHE) for typically developing children at each stage. 

This collection method categorizes children’s performance in every vital classification: Physical Examination, Language, Cognitive, Social-Emotional, Literacy and Mathematics.

We plan on researching how COVID-19 impacted different groups of students. We will analyze these groups in two categories: based on their age or based on whether they have an Individualized Educational Program (IEP) or not.

The dataset used for this analysis contains 180 observations and 10 variables in the main structured table (prof_2018_2019). These variables are `Category`, `WHE Bottom`, `WHE Top`, `# Children`, `Average`, `# Below`, `% Below`, `# Meeting / Exceeding`, `% Meeting / Exceeding and Time Period.

For analysis. We did not analyze WHE Bottom, WHE Top, or Average, as we used them only as benchmarks.

## 2. Data

We are currently tidying up the 6 datasets we plan on using for our research. Each of them presents specific data collected in a school year (e.g. 2018-2019, 2023-2024).

Meanwhile, since the datasets from the NGO have the same variables, we will use a past dataset to explain the structure. Our variables are: 
- `Bottom`: Lowest Score (for agency/ classroom)
- `Top`: Highest Score (for agency/ classroom
- `# Children`: Total # of children assessed
- `Average` : Average Score of children (for agency/ classroom)
- `# Below` :# of children below developmental expectations
- `% Below` : % of children below developmental expectations

```{r reading-data}
prof_2018_2019 <- read_excel("../data/ignore/file_show.xlsx")
```

```{r visualizing-data}
glimpse(prof_2018_2019)
```

```{r skimming-data}
skim(prof_2018_2019)
```

## 3. Data analysis plan

We will analyze our data from the following perspective:
- Pre-COVID-19: `2018_2019`
- COVID-19: `2020_2021` and `2021_2022`
- Post-COVID-19: `2022_2023` , `2023_2024` and `2024_2025`

For our research questions, we will analyze the overall performance in each general academic category: `Cognitive`, `Language`, `Literacy`, `Mathematics`, `Physical`, and `Social-Emotional`. 

For an in-depth categorization of our data, we will divide the data based on `IEP` and `Age`. 

To compare student success between these groups, we decided to create a new variable representing the percentage of students who meet or exceed expectations. We will analyze the data based on 3 timelines (Spring, Winter, and Fall) within a school year. 

#### Graphs:

1. Ridgeline Density Plot - General:  We aim to show how the pandemic affected students overall. To visualize this, we plan to use a density ridgeline graph, where the x-axis represents the Season + Year and the y-axis represents the percentage of children meeting or exceeding expectations. We also plan to separate the data by category using facets for clearer comparisons.

2. Facet Line Graph - Age Groups: We will specifically focus on how the COVID-19 pandemic impacted different age groups, ranging from 0 to 5. We plan to use a line graph, where the x-axis represents the Season + Year and the y-axis represents the percentage of students who meet or exceed expectations. The graph will be faceted by category, and different line colors will represent each age group.

3. Facet Density Plot - IEP Groups: We will use similar axes (x = Season + Year, y = percentage of students who meet or exceed expectations) and the same facet categories as in our second graph. However, instead of focusing on different age groups, this graph will compare students with IEPs to those without IEPs to determine whether children with IEPs were more affected by the pandemic. For now, this graph will include data only up to Winter 2023, since we do not yet have IEP data for the 2023–2024 and 2024–2025 academic years. We are currently waiting for that information from Monica.

#### Preliminary exploratory data analysis

##### Pre-Covid Graph
With this graph, we aim to understand how successful Promise Early Education was during a pre-COVID academic year (2018-2019) by comparing the percentage of students who met or exceeded expectations across the fall, winter, and spring assessments.
The x-axis represents the Fall, Winter, and Spring assessments for the 2018–2019 academic year, while the y-axis shows the percentage of students who met or exceeded expectations. The graph includes six facets, each representing a different learning domain: `Cognitive`, `Language`, `Physical Development`, `Social-Emotional`, `Literacy`, and `Mathematics`.

```{r categorized-data}
categorized_data_2018_2019 <- prof_2018_2019 |>
  group_by(`Time Period`, `Category`) |>
  summarise(
    `Total # Children` = sum(`# Children`, na.rm = TRUE), 
    `Meeting / Exceeding Children` = sum(`# Meeting / Exceeding`, na.rm = TRUE), 
    `% percentage` = round(`Meeting / Exceeding Children` / `Total # Children` * 100, 2),
    .groups = "drop"
  )
```

```{r mapping-categorized-table}
ggplot(categorized_data_2018_2019, aes(x = `Time Period`, y = `% percentage`)) +
  geom_col() +
  facet_wrap(~ `Category`) +
  ylim(0, 100) +
  labs(
    title = "Percentage of Children Meeting/Exceeding Expectations",
    subtitle = "by Time Period and Category",
    x = "Time Period",
    y = "Percentage (%)"
  ) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Thanks to this graph, we can see how successful Promise Early Education was in categories like `Physical`. Moreover, we can see the growth of the students from the beggining (Fall) to the end of the school year (Winter and Sping), especially in categories like `Cognitive` and `Social-Emotional` where the percentage grew from around 50% to over 80%. 

##### Facet Line Graph
## First read the data
```{r reading-cleaned-data}
X2018_2019 <- read_excel("../data/ignore/2018-2019.xlsx")
X2020_2021 <- read_excel("../data/ignore/2020-2021.xlsx")
X2021_2022 <- read_excel("../data/ignore/2021-2022.xlsx")
X2022_2023 <- read_excel("../data/ignore/2022-2023.xlsx")
X2023_2024 <- read_excel("../data/ignore/2023-2024.xlsx")
X2024_2025 <- read_excel("../data/ignore/2024-2025.xlsx")

```


```{r fig.alt = "This faceted line chart tracks the percentage of children meeting learning goals across six specific developmental categories: Cognitive, Language, Literacy, Mathematics, Physical, and Social-Emotional. The data covers the 2018–2019 to 2024–2025 school years and organizes time by season: Fall (F), Winter (W) and Spring (S) to highlight progress throughout each year. Each colored line represents a different age group: Dark Blue is 0–1 years, Purple is 1–2 years, Pink is 2–3 years, Orange is 3–4 years and Yellow is 4–5 years. The purpose of this chart is to visualize how the COVID-19 pandemic disrupted learning in these specific areas and to track how children's development has recovered over time."}

# Merge data
data <- bind_rows( X2018_2019 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), year = "18-19"),
X2020_2021 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), year = "20-21"), 
X2021_2022 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), year = "21-22"),
X2022_2023 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), year = "22-23"),
X2023_2024 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), year = "23-24"),
X2024_2025 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), year = "24-25")
) |>
  mutate(
season = case_when(
str_detect(`Time Period`, "Fall") ~ "F",
str_detect(`Time Period`, "Winter") ~ "W",
str_detect(`Time Period`, "Spring") ~ "S"
    ),
label = paste0(season, year),
sort_num = as.numeric(str_sub(year, 1, 2)) * 10 + match(season, c("F","W","S"))
  ) |>
drop_na() |>
group_by(label, Category, Age, sort_num) |>
summarise(percent = sum(`# Meeting / Exceeding`) / sum(`# Children`) * 100, .groups = "drop")

# I will make the plot and save it to covid_plot variable
covid_plot <- ggplot(data, aes(reorder(label, sort_num), percent, color = Age, group = Age)) +
geom_line(size = 1.5, alpha = 0.8) +
geom_point(size = 2.5, alpha = 0.9) +
facet_wrap(~Category) +
scale_color_viridis_d(option = "plasma") +
ylim(0, 100) +
labs(title = "COVID-19 Impact on Age Groups", 
x = "Time Period (F=Fall, W=Winter, S=Spring)", 
y = "Meeting / Exceeding Percentage", 
caption = "Data source: Promise Early Education Programs" ) +
  
theme_minimal() +
theme(
axis.text.x = element_text(angle = 90, size = 9),
strip.text = element_text(face = "bold"),
legend.position = "bottom"
  )

#Save
ggsave("covid_clear_plot.png", covid_plot, width = 20, height = 12, dpi = 300)
covid_plot
```

As we see COVID-19 pandemic made it much harder for young children to learn and grow, especially in Social-Emotional and Language skills. Children stayed physically active and healthy but they struggled more with managing feelings and talking to others. The data reveals that the youngest children aged 0–2 have had the biggest drops in progress recently and that happens likely because they were born during stressful times with less social interaction. While they recover well during the school year, this data still highlights a strong need for summer bridge programs and focused social-emotional support to help them stay more on track and grow.

```{r}


# First Mix all the data from different years
data <- bind_rows( X2018_2019 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), 
           `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), 
year = "18-19"),
X2020_2021 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), 
           `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), 
 year = "20-21"), 
X2021_2022 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), 
           `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), 
year = "21-22"),
X2022_2023 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), 
           `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), 
year = "22-23"),
X2023_2024 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), 
           `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), 
year = "23-24"),
X2024_2025 |> select(Category, `# Children`, Age, `# Meeting / Exceeding`, `Time Period`) |> 
mutate(`# Children` = suppressWarnings(as.numeric(`# Children`)), 
           `# Meeting / Exceeding` = suppressWarnings(as.numeric(`# Meeting / Exceeding`)), 
year = "24-25")
)

## add labels and calculate percentages
data <- data |> mutate(
season = case_when(
str_detect(`Time Period`, "Fall") ~ "F",
str_detect(`Time Period`, "Winter") ~ "W",
str_detect(`Time Period`, "Spring") ~ "S"
    ),
label = paste0(season, year),
sort_num = as.numeric(str_sub(year, 1, 2)) * 10 + match(season, c("F","W","S"))
  ) |>
drop_na() |>
group_by(label, Category, Age, sort_num) |>
summarise(percent = sum(`# Meeting / Exceeding`) / sum(`# Children`) * 100, .groups = "drop")

# Show where COVID start and where it ends
covid_start <- 183.5
covid_end <- 203.5

# Create the plot
covid_plot <- ggplot(data, aes(x = sort_num, y = percent, color = Age, group = Age)) +
annotate("rect", xmin = covid_start, xmax = covid_end, ymin = 0, ymax = 100, 
fill = "#ff4443", alpha = 0.15) +
geom_vline(xintercept = covid_start, color = "#cc0000", size = 1.5, 
linetype = "solid", alpha = 0.8) +
geom_vline(xintercept = covid_end, color = "#cc0000", size = 1.5, 
linetype = "dashed", alpha = 0.6) +
geom_line(size = 1.5, alpha = 0.85) +
geom_point(size = 2.8, alpha = 0.9) +
annotate("text", x = covid_start - 5, y = 95, label = "PRE-COVID", 
color = "gray40", fontface = "bold", size = 3, alpha = 0.6) +
annotate("text", x = (covid_start + covid_end) / 2, y = 95, label = "COVID-19\nIMPACT", 
color = "#cc0000", fontface = "bold", size = 3.5, alpha = 0.7) +
annotate("text", x = covid_end + 10, y = 95, label = "RECOVERY", 
color = "gray40", fontface = "bold", size = 3, alpha = 0.6) +
facet_wrap(~Category, ncol = 3) +
scale_color_viridis_d(option = "plasma", 
labels = c("0-1 yr", "1-2 yrs", "2-3 yrs", "3-4 yrs", "4-5 yrs")) +
scale_x_continuous(breaks = unique(data$sort_num),
labels = unique(data$label)[order(unique(data$sort_num))]) +
ylim(0, 100) +
labs(title = "COVID-19 Impact on Child Development Across Age Groups",
subtitle = "Red zone indicates COVID-19 impact period (2019-2021) | Lines show recovery trajectory",
x = "Time Period (F=Fall, W=Winter, S=Spring)", 
y = "Meeting / Exceeding Learning Goals (%)", 
color = "Age Group",
caption = "Data source: Promise Early Education Programs") +
theme_minimal(base_size = 12) +
theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40"),
axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
strip.text = element_text(face = "bold", size = 11),
legend.position = "bottom",
legend.title = element_text(face = "bold"),
panel.grid.minor = element_blank(),
panel.grid.major.x = element_line(color = "gray100", size = 0.3),
plot.caption = element_text(hjust = 1, color = "gray60", size = 9))

# Save and show
ggsave("covid_impact_enhanced.png", covid_plot, width = 20, height = 12, dpi = 300)
covid_plot
```

