---
title: "README"
author: "Gold Analysts - Mekkawy, Nissan, Razvan"
output: github_document
---

# data

Data file(s) placed in this folder.

```{r reading-cleaned-data}
library(readxl)
library(dplyr)
X2018_2019 <- read_excel("../data/ignore/2018-2019.xlsx")
X2020_2021 <- read_excel("../data/ignore/2020-2021.xlsx")
X2021_2022 <- read_excel("../data/ignore/2021-2022.xlsx")
X2022_2023 <- read_excel("../data/ignore/2022-2023.xlsx")
X2023_2024 <- read_excel("../data/ignore/2023-2024.xlsx")
X2024_2025 <- read_excel("../data/ignore/2024-2025.xlsx")
```

## 2018-2019

- `Category`: 6 categories analyzed: Social-Emotional, Physical, Language, Cognitive, Literacy, Mathematics
- `# Children`: Total # of children assessed in each age group
- `Age`: 5 age groups analyzed: 0-1, 1-2, 2-3, 3-4, 4-5
- `IEP`: IEP status: N/A (0-1, 1-2, 2-3 and some 3-4, 4-5 depending on the year), Yes or No (most 3-4 and 4-5)
- `# Meeting/Exceeding` : # of children meeting or exceeding expectations
- `% Meeting/Exceeding` : % of children meeting or exceeding expectations
- `Time Period` : trimester and year analyzed (Fall 2018-2019, Winter 2018-2019, Spring 2018-2019)

```{r visualizing-data-1}
ls(X2018_2019)
glimpse(X2018_2019)
```

## 2020-2021

- `Category`: 6 categories analyzed: Social-Emotional, Physical, Language, Cognitive, Literacy, Mathematics
- `# Children`: Total # of children assessed in each age group
- `Age`: 5 age groups analyzed: 0-1, 1-2, 2-3, 3-4, 4-5
- `IEP`: IEP status: N/A (0-1, 1-2, 2-3 and some 3-4, 4-5 depending on the year), Yes or No (most 3-4 and 4-5)
- `# Meeting/Exceeding` : # of children meeting or exceeding expectations
- `% Meeting/Exceeding` : % of children meeting or exceeding expectations
- `Time Period` : trimester and year analyzed (Fall 2020-2021, Winter 2020-2021, Spring 2020-2021)

```{r visualizing-data-2}
ls(X2020_2021)
glimpse(X2020_2021)
```

## 2021-2022

- `Category`: 6 categories analyzed: Social-Emotional, Physical, Language, Cognitive, Literacy, Mathematics
- `# Children`: Total # of children assessed in each age group
- `Age`: 5 age groups analyzed: 0-1, 1-2, 2-3, 3-4, 4-5
- `IEP`: IEP status: N/A (0-1, 1-2, 2-3 and some 3-4, 4-5 depending on the year), Yes or No (most 3-4 and 4-5)
- `# Meeting/Exceeding` : # of children meeting or exceeding expectations
- `% Meeting/Exceeding` : % of children meeting or exceeding expectations
- `Time Period` : trimester and year analyzed (Fall 2021-2022, Winter 2021-2022, Spring 2021-2022)

```{r visualizing-data-3}
ls(X2021_2022)
glimpse(X2021_2022)
```

## 2022-2023

- `Category`: 6 categories analyzed: Social-Emotional, Physical, Language, Cognitive, Literacy, Mathematics
- `# Children`: Total # of children assessed in each age group
- `Age`: 5 age groups analyzed: 0-1, 1-2, 2-3, 3-4, 4-5
- `IEP`: IEP status: N/A (0-1, 1-2, 2-3 and some 3-4, 4-5 depending on the year), Yes or No (most 3-4 and 4-5)
- `# Meeting/Exceeding` : # of children meeting or exceeding expectations
- `% Meeting/Exceeding` : % of children meeting or exceeding expectations
- `Time Period` : trimester and year analyzed (Fall 2022-2023, Winter 2022-2023, Spring 2022-2023)

```{r visualizing-data-4}
ls(X2022_2023)
glimpse(X2022_2023)
```

## 2023-2024

- `Category`: 6 categories analyzed: Social-Emotional, Physical, Language, Cognitive, Literacy, Mathematics
- `# Children`: Total # of children assessed in each age group
- `Age`: 5 age groups analyzed: 0-1, 1-2, 2-3, 3-4, 4-5
- `IEP`: IEP status: N/A (0-1, 1-2, 2-3 and some 3-4, 4-5 depending on the year), Yes or No (most 3-4 and 4-5)
- `# Meeting/Exceeding` : # of children meeting or exceeding expectations
- `% Meeting/Exceeding` : % of children meeting or exceeding expectations
- `Time Period` : trimester and year analyzed (Fall 2023-2024, Winter 2023-2024, Spring 2023-2024)

```{r visualizing-data-5}
ls(X2023_2024)
glimpse(X2023_2024)
```

## 2024-2025

- `Category`: 6 categories analyzed: Social-Emotional, Physical, Language, Cognitive, Literacy, Mathematics
- `# Children`: Total # of children assessed in each age group
- `Age`: 5 age groups analyzed: 0-1, 1-2, 2-3, 3-4, 4-5
- `IEP`: IEP status: N/A (0-1, 1-2, 2-3 and some 3-4, 4-5 depending on the year), Yes or No (most 3-4 and 4-5)
- `# Meeting/Exceeding` : # of children meeting or exceeding expectations
- `% Meeting/Exceeding` : % of children meeting or exceeding expectations
- `Time Period` : trimester and year analyzed (Fall 2024-2025, Winter 2024-2025, Spring 2024-2025)

```{r visualizing-data-6}
ls(X2024_2025)
glimpse(X2024_2025)
```